---
title: "Fitting Classification Trees"
author: "Valentin Savchenko"
date: "1/23/2021"
output: html_document
---

```{r, setoptions, echo = F, warning = F}
library(knitr)
opts_chunk$set(echo = T)
opts_chunk$set(warning = F)
opts_chunk$set(message = F)
set.seed(3)

library(tree)
library(dplyr)
library(tibble)
```

Load the Carseats dataset and preview its content:
```{r}
library(ISLR)
attach(Carseats)
head(Carseats)
```

Introduce a binary variable High, which signals whether sales considered high: 
```{r}
carseats <- as_tibble(Carseats) %>% mutate(High=as.factor(Sales>8))
head(carseats)
```

Split available data into training and test sets:
```{r}
trainSize <- 0.5 * nrow(carseats)
trainSize
trainIdx <- sample(1:nrow(carseats), trainSize)
testCarseats <- carseats[-trainIdx, ]
```

Fit a decision tree to predict whether sales are high based on all original predictors:
```{r}
model <- tree(High ~.- Sales, carseats, subset=trainIdx)
summary(model)
plot(model)
text(model, pretty=0)
```

Evaluate quality of the tree on the test set:
```{r}
predictions <- predict(model, testCarseats, type="class")
table(predictions, testCarseats$High)
correctPredictions <- sum(predictions == testCarseats$High) / nrow(testCarseats)
correctPredictions
```

Attempt to improve quality of the model. Perform cross-validation in order to determine the optimal level of tree complexity. Guide the cross-validation and prunning process by the classification error rate:
```{r}
crossValidation <- cv.tree(model, FUN=prune.misclass)
names(crossValidation)
```

Determine the number of terminal nodes that results in the lowest cross-validation error rate:
```{r}
crossValidation$dev
optimalSizeId <- which.min(crossValidation$dev)

crossValidation$size
optimalSize <- crossValidation$size[optimalSizeId]
optimalSize
```

Plot the error rate as a function of both the number of terminal nodes of each tree considered and the value of the cost-complexity parameter used:
```{r}
par(mfrow =c(1,2))
plot(crossValidation$size, crossValidation$dev, type="b")
plot(crossValidation$k, crossValidation$dev, type="b")
```

Prune the original tree:
```{r}
pruneModel <- prune.misclass(model, best=optimalSize)
summary(pruneModel)
plot(pruneModel)
text(pruneModel, pretty=0)
```

Evaluate performance of the pruned tree:
```{r}
prunePredictions <- predict(pruneModel, testCarseats, type="class")
table(prunePredictions, testCarseats$High)
correctPrunePredictions <- sum(prunePredictions == testCarseats$High) / nrow(testCarseats)
correctPrunePredictions
```